---
author: Xinlei Huang (xhuang84), Jaccob Mau (jcmau2), Mohammad Khan (mfkahn5), Jonathan Temkin (jtemkin3)
output:
  html_document:
    theme: readable
    toc: true
  pdf_document: default
title: Data Analysis Project - Proposal
urlcolor: cyan
editor_options: 
  markdown: 
    wrap: 72
---

# Topic

This project will research factors that might predict life expectancy across multiple
regions globally.

# Description of the data set

For this project we propose using the `Life Expectancy (WHO)` data set
which is available
[here](https://www.kaggle.com/datasets/kumarajarshi/life-expectancy-who).
The data is an amalgamation of two sources. The primary source is The
Global Health Observatory (GHO) repository, which is maintained by the
World Health Organization (WHO) for the purpose of tracking demographic
and health factors impacting mortality. The data spans 193 countries and
15 years, from 2000 to 2015. This dataset is supplemented with United
Nations economic data. The resulting dataset will enable us to assess
the impact of immunization, economic and social factors on mortality.

The dataset was compiled by Deeksha Russel and Duan Wang. It is a
curated subset of the Global Health Observatory database, with only
those variables deemed most critical and representative being retained in
the file dataset. The dataset has also been cleaned to remove countries
for which data was incomplete.

```{r}
library(readr)
data = read_csv('./data/life_expectancy.csv', show_col_types = FALSE)
head(data)
```

There are 22 columns and 2938 records in the CSV file. `Life expectancy`
will be used as the response variable, and the rest are predictor
candidates. The table below lists out some important ones,

```{r}
library(knitr)
predictors = data.frame(names = c("country", "status", "adult mortality", "infant deaths", "Alcohol", "BMI", "GDP", "Population", "Schooling"),
                        types = c("categorical", "dummy", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric"))
kable(predictors)
```

# Why

It is generally accepted that developed countries have lower mortality
and higher lifespan than their less-developed peers. But how true is
this assumption? And if these countries do indeed have lower mortality,
which factors are most responsible?

Developed countries tend to have more resources, better nutrition
standards and access to advanced medical infrastructure, all of which
facilitate longevity. On the other hand, these countries may be more
likely to suffer from higher rates of obesity or alcoholism, which can
counteract those benefits. Moreover, the complex economies upon which
these countries' development is predicated may result in more stressful
lifestyles for their inhabitants, further harming mortality.

Buddhist monks live in Nepal don't spend their sleeping time on swiping
phones, breath car exhaust, or eat packaged food full of chemicals. It's
true Nepal's GDP is far behind US, but do US residents live longer than
Nepal people necessarily? Hopefully data analysis with this data set can
shed some light.
