---
Author:  RandomM53
output:
  html_document:
    theme: readable
    toc: true
  pdf_document: default
title: Factors Impacting Life Expectancy
urlcolor: cyan
editor_options: 
  markdown: 
    wrap: 72
---

# Introduction

This project will research factors that might predict life expectancy
across multiple regions globally. It is generally accepted that developed
countries have lower mortality and higher lifespan than their less-developed peers.
But how true is this assumption? And if these countries do indeed have lower mortality,
which factors are most responsible?

Developed countries tend to have more resources, better nutrition
standards and access to advanced medical infrastructure, all of which
facilitate longevity. On the other hand, these countries may be more
likely to suffer from higher rates of obesity or alcoholism, which can
counteract those benefits. Moreover, the complex economies upon which
these countries' development is predicated may result in more stressful
lifestyles for their inhabitants, further harming mortality.

Buddhist monks live in Nepal don't spend their sleeping time on swiping
phones, breath car exhaust, or eat packaged food full of chemicals. It's
true Nepal's GDP is far behind US, but do US residents live longer than
Nepal people necessarily? Hopefully data analysis with this data set can
shed some light.

## Description of the data set

For this project we propose using the `Life Expectancy (WHO)` data set
which is available
[here](https://www.kaggle.com/datasets/kumarajarshi/life-expectancy-who).
The data is an amalgamation of two sources. The primary source is The
Global Health Observatory (GHO) repository, which is maintained by the
World Health Organization (WHO) for the purpose of tracking demographic
and health factors impacting mortality. The data spans 193 countries and
15 years, from 2000 to 2015. This dataset is supplemented with United
Nations economic data. The resulting dataset will enable us to assess
the impact of immunization, economic and social factors on mortality.

The dataset was compiled by Deeksha Russel and Duan Wang. It is a
curated subset of the Global Health Observatory database, with only
those variables deemed most critical and representative being retained
in the file dataset. The dataset has also been cleaned to remove
countries for which data was incomplete.

```{r}
library(readr)
data = read_csv('./data/life_expectancy.csv', show_col_types = FALSE)
head(data)
```

There are 22 columns and 2938 records in the CSV file. `Life expectancy`
will be used as the response variable, and the rest are predictor
candidates. The table below lists out some important ones,

```{r}
library(knitr)
predictors = data.frame(names = c("country", "status", "adult mortality", "infant deaths", "Alcohol", "BMI", "GDP", "Population", "Schooling"),
                        types = c("categorical", "dummy", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric"))
kable(predictors)
```

## Column definitions

| Name | Definition |
|----|----|
| country | country |
| year | year |
| status | Developed or Developing |
| life expectancy | life expectancy in age |
| adult mortality | Adult Mortality Rates of both sexes (probability of dying between 15 and 60 years per 1000 population) |
| infant deaths | Number of Infant Deaths per 1000 population |
| alcohol | Alcohol, recorded per capita (15+) consumption (in litres of pure alcohol) |
| percentage expenditure | Expenditure on health as a percentage of Gross Domestic Product per capita (%) |
| Hepatitis B | Hepatitis B (HepB) immunization coverage among 1-year-olds (%) |
| Measles | Measles - number of reported cases per 1000 population |
| BMI | Average Body Mass Index of entire population |
| under-five deaths | Number of under-five deaths per 1000 population |
| Polio | Polio (Pol3) immunization coverage among 1-year-old (%) |
| total expenditure | General government expenditure on health as a percentage of total government expenditure (%) |
| Diphtheria | Diphtheria tetanus toxoid and pertussis (DTP3) immunization coverage among 1-year-olds (%) |
| HIV/AIDS | Deaths per 1 000 live births HIV/AIDS (0-4 years) |
| GDP | Gross Domestic Product per capita (in USD) |
| Population | Population of the country |
| thinness 1-19 years | Prevalence of thinness among children and adolescents for Age 10 to 19 (% ) |
| thinness 5-9 years | Prevalence of thinness among children for Age 5 to 9(%) |
| income composition | Human Development Index in terms of income composition of resources (index ranging from 0 to 1) |
| schooling | Number of years of Schooling(years) |

# Methods
## Data Cleaning
Original variable names are not r-friendly. We will rename them to follow snake case for easy access on further analysis.
- rename variables
- convert category variables to factors
- filter records based on column values
- omit NA values

```{r}
names(data)[names(data) == "Country"] = "country"
names(data)[names(data) == "Year"] = "year"
names(data)[names(data) == "Status"] = "status"
names(data)[names(data) == "Life expectancy"] = "life_expectancy"
names(data)[names(data) == "Adult Mortality"] = "adult_mortality"
names(data)[names(data) == "infant deaths"] = "infant_deaths"
names(data)[names(data) == "Alcohol"] = "alcohol"
names(data)[names(data) == "percentage expenditure"] = "percent_exp"
names(data)[names(data) == "Hepatitis B"] = "hepatitis"
names(data)[names(data) == "measles"] = "measles"
names(data)[names(data) == "under-five deaths"] = "under_5_deaths"
names(data)[names(data) == "Polio"] = "polio"
names(data)[names(data) == "Total expenditure"] = "total_exp"
names(data)[names(data) == "Diphtheria"] = "diphtheria"
names(data)[names(data) == "HIV/AIDS"] = "hiv_aids"
names(data)[names(data) == "Population"] = "population"
names(data)[names(data) == "thinness  1-19 years"] = "thinness_1_19"
names(data)[names(data) == "thinness 5-9 years"] = "thinness_5_9"
names(data)[names(data) == "Income composition of resources"] = "HBI"
names(data)[names(data) == "Schooling"] = "schooling"

data$country = as.factor(data$country)
data$year = as.factor(data$year)
data$status = as.factor(data$status)

data = subset(data, percent_exp > 0)
data = subset(data, hiv_aids > 0)
data = subset(data, GDP > 0)
data = na.omit(data)
```

## Identify Potential Transformations and Collinearity
Plot a matrix of scatterplots to check shapes of curves for
- `life_expectancy vs numeric predictors`
- $Predictor_x$ vs $Predictor_y$

```{r}
data_numeric_only = subset(data, select = -c(country, year, status))
pairs(data_numeric_only)
```

From above matrix of scatterplots, we can observe
- collinearity
  - `infant_death` and `under_5_death`
  - `thinness_5_9` and `thinness_1_19`
- predictors with non-linear shape of curve against life expectancy
  - `percent_exp`
  - `hiv_aids`
  - `GDP`

Take logarithm for predictors with non-linear shape of curve

```{r}
data_transformed = data.frame(data)
data_transformed["log_percent_exp"] = log(data_transformed$percent_exp)
data_transformed["log_hiv_aids"] = log(data_transformed$hiv_aids)
data_transformed["log_GDP"] = log(data_transformed$GDP)
```

Remove linearly correlated column and two category columns `country` and `year`.
- remove `thinness_5_9` and `infant_death` because they are subsets to `thiness_1_19` and `under_5_deaths` respectively
- remove `year` to get rid of time series characteristic
- remove `country` to mitigate geographic influence
- omit NA values

```{r}
data_cleaned = subset(data_transformed, select = -c(country, year, percent_exp, hiv_aids, GDP, thinness_5_9, infant_deaths))
data_cleaned = na.omit(data_cleaned)
```

## Model Selection
### compare additive models with predictors and transformed predictors

```{r}
fit_add_orign = lm(life_expectancy ~ ., data = subset(data, select = -c(country, year)))
summary(fit_add_orign)
fit_add_trans = lm(life_expectancy ~ ., data = data_cleaned)
summary(fit_add_trans)
```
Both models are significant under 0.01 significant level, and the one with transformed data has higher `r-squared` value. Further optimization will be based the data set `data_cleaned`.

### perform a regression with level 2 interaction

```{r}
fit_int = lm(life_expectancy ~ .^2, data = data_cleaned)
```

### compare the interaction model with additive model

```{r}
anova(fit_add_trans, fit_int)
```
`p-value` is small enough to reject the hypothesis that interactive predictors were in-significant, hence interactive model `fit_int` is preferred over the additive one.

### backward step-wise selection with AIC for both the additive and interactive model

```{r}
fit_add_selected = step(fit_add_trans, direction = "backward", trace = 0)
summary(fit_add_selected)

fit_selected = step(fit_int, direction = "backward", trace = 0)
summary(fit_selected)
```

Noticed that not all selected predictors are significant with a high significance level such as 99%.
Try to fit a simple model which excludes predictors with `p-value > 0.01`

```{r}
df_coef = data.frame(summary(fit_selected)$coefficients)
row.names(df_coef)[df_coef["Pr...t.."] < 0.01]
(predictor_formula_str = paste(row.names(df_coef)[df_coef["Pr...t.."] < 0.01][-1], collapse = " + "))
fit_small = lm(life_expectancy ~ adult_mortality + Measles + BMI + polio + total_exp + population + thinness_1_19 + HBI + log_hiv_aids + status:adult_mortality + status:hepatitis + status:total_exp + status:thinness_1_19 + status:HBI + status:schooling + adult_mortality:BMI + adult_mortality:schooling + adult_mortality:log_hiv_aids + adult_mortality:log_GDP + alcohol:hepatitis + alcohol:Measles + alcohol:log_percent_exp + alcohol:log_hiv_aids + hepatitis:total_exp + hepatitis:log_percent_exp + BMI:population + BMI:thinness_1_19 + BMI:schooling + under_5_deaths:polio + under_5_deaths:HBI + under_5_deaths:schooling + polio:HBI + total_exp:thinness_1_19 + total_exp:HBI + diphtheria:HBI + diphtheria:schooling + thinness_1_19:HBI + thinness_1_19:schooling + thinness_1_19:log_hiv_aids + HBI:schooling + HBI:log_hiv_aids + schooling:log_percent_exp + schooling:log_GDP + log_percent_exp:log_GDP,
               data = data_cleaned)
anova(fit_small, fit_selected)
```

`fit_selected` is preferred over the smaller model on anova analysis, but how about the Test-Train Split?

```{r, warning=FALSE}
set.seed(233)
total = nrow(data_cleaned)
trn_idx = sample(total, floor(total * 0.8))
data_trn = data_cleaned[trn_idx, ]
data_tst = data_cleaned[-trn_idx, ]

rmse <- function(model, new_data) {
  y_hat = predict(model, newdata = new_data)
  return(sqrt(mean((y_hat - data_tst$life_expectancy)^2)))
}

fit_complex = lm(life_expectancy ~ status + adult_mortality + alcohol + hepatitis + 
    Measles + BMI + under_5_deaths + polio + total_exp + diphtheria + 
    population + thinness_1_19 + HBI + schooling + log_percent_exp + 
    log_hiv_aids + log_GDP + status:adult_mortality + status:alcohol + 
    status:hepatitis + status:BMI + status:total_exp + status:thinness_1_19 + 
    status:HBI + status:schooling + adult_mortality:BMI + adult_mortality:population + 
    adult_mortality:schooling + adult_mortality:log_hiv_aids + 
    adult_mortality:log_GDP + alcohol:hepatitis + alcohol:Measles + 
    alcohol:BMI + alcohol:population + alcohol:HBI + alcohol:schooling + 
    alcohol:log_percent_exp + alcohol:log_hiv_aids + hepatitis:Measles + 
    hepatitis:total_exp + hepatitis:population + hepatitis:HBI + 
    hepatitis:log_percent_exp + Measles:BMI + Measles:under_5_deaths + 
    Measles:total_exp + Measles:HBI + Measles:log_percent_exp + 
    Measles:log_hiv_aids + Measles:log_GDP + BMI:under_5_deaths + 
    BMI:total_exp + BMI:diphtheria + BMI:population + BMI:thinness_1_19 + 
    BMI:HBI + BMI:schooling + under_5_deaths:polio + under_5_deaths:total_exp + 
    under_5_deaths:population + under_5_deaths:HBI + under_5_deaths:schooling + 
    under_5_deaths:log_percent_exp + under_5_deaths:log_hiv_aids + 
    under_5_deaths:log_GDP + polio:total_exp + polio:population + 
    polio:HBI + total_exp:diphtheria + total_exp:thinness_1_19 + 
    total_exp:HBI + total_exp:log_percent_exp + total_exp:log_GDP + 
    diphtheria:HBI + diphtheria:schooling + diphtheria:log_percent_exp + 
    population:HBI + population:log_percent_exp + population:log_hiv_aids + 
    population:log_GDP + thinness_1_19:HBI + thinness_1_19:schooling + 
    thinness_1_19:log_hiv_aids + thinness_1_19:log_GDP + HBI:schooling + 
    HBI:log_hiv_aids + schooling:log_percent_exp + schooling:log_hiv_aids + 
    schooling:log_GDP + log_percent_exp:log_GDP,
    data = data_trn)
fit_simple = lm(life_expectancy ~ adult_mortality + Measles + BMI + polio + total_exp + 
    population + thinness_1_19 + HBI + log_hiv_aids + status:adult_mortality + 
    status:hepatitis + status:total_exp + status:thinness_1_19 + 
    status:HBI + status:schooling + adult_mortality:BMI + adult_mortality:schooling + 
    adult_mortality:log_hiv_aids + adult_mortality:log_GDP + 
    alcohol:hepatitis + alcohol:Measles + alcohol:log_percent_exp + 
    alcohol:log_hiv_aids + hepatitis:total_exp + hepatitis:log_percent_exp + 
    BMI:population + BMI:thinness_1_19 + BMI:schooling + under_5_deaths:polio + 
    under_5_deaths:HBI + under_5_deaths:schooling + polio:HBI + 
    total_exp:thinness_1_19 + total_exp:HBI + diphtheria:HBI + 
    diphtheria:schooling + thinness_1_19:HBI + thinness_1_19:schooling + 
    thinness_1_19:log_hiv_aids + HBI:schooling + HBI:log_hiv_aids + 
    schooling:log_percent_exp + schooling:log_GDP + log_percent_exp:log_GDP,
    data = data_trn)
trn_rmse_complex = rmse(fit_complex, data_trn)
tst_rmse_complex = rmse(fit_complex, data_tst)
trn_rmse_simple = rmse(fit_simple, data_trn)
tst_rmse_simple = rmse(fit_simple, data_tst)

library(knitr)
kable(data.frame(model = c("Simple", "Complex"),
                 "Train RMSE" = c(trn_rmse_simple, trn_rmse_complex),
                 "Test RMSE" = c(tst_rmse_simple, tst_rmse_complex)))
```

The simple model has lower RMSE on both train and test data sets. Simple model will be selected on further analysis.

Lets compare the Simple model to the "really simple" additive model that we originally started out with.

```{r}

# Create a model based on the training dataset with the predictors chosen via backward AIC
fit_add = lm(life_expectancy ~ status + adult_mortality + hepatitis + BMI + under_5_deaths + total_exp + diphtheria + population + thinness_1_19 + HBI + schooling + log_percent_exp + log_hiv_aids + log_GDP, data = data_trn)


trn_rmse_add = rmse(fit_add, data_trn)
tst_rmse_add = rmse(fit_add, data_tst)

kable(data.frame(model = c("Simple", "Additive"),
                 "Train RMSE" = c(trn_rmse_simple, trn_rmse_add),
                 "Test RMSE" = c(tst_rmse_simple, tst_rmse_add)))   
```

Interestingly, the additive model has lower train RMSE than the simple interaction model.
However, the test RMSE is way lower for the simple interaction model and thus we continue to prefer the simple interaction model.


## Colinearity Analysis
#### @Jonathon

## Model Diagnostics
#### @Jaccob

```{r}
shapiro.test(resid(fit_small))
qqnorm(resid(fit_small), col = "darkgrey")
qqline(resid(fit_small), col = "dodgerblue", lwd = 2)


shapiro.test(resid(fit_add_selected))
qqnorm(resid(fit_add_selected), col = "darkgrey")
qqline(resid(fit_add_selected), col = "dodgerblue", lwd = 2)
```


## Outlier Diagnostics
#### @Fahad

The Q-Q plot definitely makes it seem that the normality assumptions is violated.
Lets try to remove outliers and see if we can improve the normality of the residuals.

To check for outlier data points having an outside influence on our model, we will check the cooks distance of the data in our dataset.

```{r}
# calculating cooks distance for all the data points for the selected simple model that was trained on all the data
small_int_mod_cd = cooks.distance(fit_small)
sum(small_int_mod_cd > 4/ length(small_int_mod_cd))

# calculating cooks distance for all the data points for the selected additive model that was trained on all the data
add_mod_cd = cooks.distance(fit_add_selected)
sum(add_mod_cd > 4/ length(add_mod_cd))
```

There are 121 outliers that are having a high influence on our interaction model and 125 data points that are having a high influence on our additive model.

Lets remove these and run the diagnostics again.

```{r}
fit_small_int_fix = lm(life_expectancy ~ adult_mortality + Measles + BMI + polio + total_exp + population + thinness_1_19 + HBI + log_hiv_aids + status:adult_mortality + status:hepatitis + status:total_exp + status:thinness_1_19 + status:HBI + status:schooling + adult_mortality:BMI + adult_mortality:schooling + adult_mortality:log_hiv_aids + adult_mortality:log_GDP + alcohol:hepatitis + alcohol:Measles + alcohol:log_percent_exp + alcohol:log_hiv_aids + hepatitis:total_exp + hepatitis:log_percent_exp + BMI:population + BMI:thinness_1_19 + BMI:schooling + under_5_deaths:polio + under_5_deaths:HBI + under_5_deaths:schooling + polio:HBI + total_exp:thinness_1_19 + total_exp:HBI + diphtheria:HBI + diphtheria:schooling + thinness_1_19:HBI + thinness_1_19:schooling + thinness_1_19:log_hiv_aids + HBI:schooling + HBI:log_hiv_aids + schooling:log_percent_exp + schooling:log_GDP + log_percent_exp:log_GDP,
               data = data_cleaned,
               subset = small_int_mod_cd < 4/ length(small_int_mod_cd))

fit_add_fix = lm(life_expectancy ~ status + adult_mortality + hepatitis + BMI + under_5_deaths + total_exp + diphtheria + population + thinness_1_19 + HBI + schooling + log_percent_exp + log_hiv_aids + log_GDP,
               data = data_cleaned,
               subset = add_mod_cd < 4/ length(add_mod_cd))


shapiro.test(resid(fit_small_int_fix))
qqnorm(resid(fit_small_int_fix), col = "darkgrey")
qqline(resid(fit_small_int_fix), col = "dodgerblue", lwd = 2)


shapiro.test(resid(fit_add_fix))
qqnorm(resid(fit_add_fix), col = "darkgrey")
qqline(resid(fit_add_fix), col = "dodgerblue", lwd = 2)
```

We managed to improve the normality of the residuals for the interactive model (the Q-Q plot looks much better) however the normality of the residuals of the additive model was not improved.

Thus we still continue to prefer the interaction model even though the Shapiro-Wilk test continues to reject for all reasonable alphas.
































